---
title: "HCHS ACE Analysis"
output: html_document
date: "2023-03-10"
author: "Anne Klimesch"
note: |
  This script combines the variables of interest of different data sets into one new dataset for analysis. The datasets that will be combined are: 
  1. SES data and ACE data ("HCHS_Data_SPSS_labelled_sozio_HCHS_SES_Index_2023_03_09.sav"), 
  2. structural subcortical MRI data ("aseg.sav"), and 
  3. structural cortical MRI data ("aparc_rh_thickness_destrieux_stats.txt",  "aparc_rh_volume_destrieux_stats.txt"). 
  
  Data are then preprocessed and prepared for analysis. The output of this code will be three preprocessed datasets: 
  1. "preimp_no_na.csv" which will be used for the main analyses, 
  2. "preimp_no_na_excl_outlier.csv" which will be used for sensitivity analyses, and 
  3. "preimp_prepped_for_imputation.csv" which will be used in the code "02_imputation.Rmd" to impute the missing data and perform further sensitivity analyses with the resulting dataset. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages
```{r, include=FALSE}
library(dplyr)
library(stringr)  #extract parts of a variable for each case of this variable; organize columns when reading the mri txt files; to remove spaces from values
library(haven)    #to import spss data files
library(foreign)
library(naniar)   #for missing variable analysis 
library(nortest)  #for missing variable analysis 
library(stats)    #for missing variable analysis 
library(kableExtra)#for missing data analysis
library(ggplot2)  #for histograms
library(gridExtra)#for presentation of histograms 
library(finalfit) #for visualization of missing data patterns 
library(psych)    #for descriptives and intercorrelation matrix
library(sjPlot)   #for plotting descriptives
library(stargazer)#for plotting descriptives
library(ggplot2)  #for plotting 
library(gridExtra)#for plotting 
library(corrplot)    #for intercorrelation matrix
library(PerformanceAnalytics) #for heatmap
library(corrplot) #for colored intercorrelation matrix
library(knitr)    #for nice_table
library(rempsyc)  #for nice_table
#library(pcaMethods) # for misscompare
library(visdat) #for visualization of missing data pattern for many variables at the same time
library(apaTables)
library("Hmisc")
library(DescTools) #for kendall b correlation coefficient 
```

IMPORTING MRI AND SES/QUESTIONNAIRE DATA
Import subcortical neuro data file, SES-data file (SES-Index output of HCHS SES study), and files with DLPFC neuro data 
```{r}
mri_data_subcort <- read_sav("C:your_path_to_data/aseg.sav")

ses_data <- read.spss("C:your_path_to_data/HCHS_Data_SPSS_labelled_sozio_HCHS_SES_Index_2023_03_18.sav", use.value.label = FALSE, to.data.frame=TRUE)

mri_data_cort_1 <- read.table("aparc_rh_thickness_destrieux_stats.txt", sep = "", check.names = TRUE, header = TRUE, dec = ".") 
colnames(mri_data_cort_1)[1] <- "pseudonym" #rename the first column to pseudonym

mri_data_cort_2 <- read.table("aparc_rh_volume_destrieux_stats.txt", sep = "", check.names = TRUE, header = TRUE, dec = ".") 
colnames(mri_data_cort_2)[1] <- "pseudonym" #rename the first column to pseudonym
```

PREPARATION CORTICAL MRI DATA
Choose subset with rDLPFC thickness data; based on Destrieux et al. (2010)
middle frontal gyrus, F2 (rh_G_front_middle_thickness)
superior frontal gyrus (rh_G_front_sup_thickness)
```{r}
mri_data_cort_complete <- merge(mri_data_cort_1, mri_data_cort_2, by = "pseudonym", all = TRUE) #merge the two files with cortical data 
mri_dlpfc_subset <- mri_data_cort_complete[, c("pseudonym", #choose variables of interest for research question
                                               "rh_G_front_middle_thickness",
                                               "rh_G_front_sup_thickness",
                                               "eTIV.x")]
```

PREPARATION OF SUBCORTICAL MRI DATA
```{r}
mri_subcort_subset <- mri_data_subcort[, c("Pseudonym_N_MRT", #choose variables of interest for research question
                                           "RightHippocampus", 
                                           "RightAmygdala", 
                                           "LeftHippocampus", 
                                           "LeftAmygdala", 
                                           "bilatHippocampus", 
                                           "bilatAmygdala",
                                           "lhCortexVol",
                                           "rhCortexVol",
                                           "TotalGrayVol",
                                           "EstimatedTotalIntraCranialVol")]

colnames(mri_subcort_subset)[1] <- "pseudonym" #rename the first column to pseudonym
mri_subcort_and_dlpfc <- merge(mri_subcort_subset, mri_dlpfc_subset, by = "pseudonym", all = TRUE) #merge df with subcortical and cortical data by pseudonym
mri_data_complete <- mri_subcort_and_dlpfc 
```
n(mri_data_complete)=2525

PREPARATION TIV VALUES
Save the text file with the TIV values as a sav file.
Rename variables 1 and 3. 
```{r}
#load dataset
tiv_data <- read.delim("tiv.txt", header = TRUE, sep = " ", dec = ".") #import .txt file with tiv values
names(tiv_data) <- c("v1", "v2", "cat12_tiv") #rename variables 

#remove missing values 
tiv_data[is.na(tiv_data$cat12_tiv), ] #check for missing tiv values 
tiv_data_subset <- tiv_data[complete.cases(tiv_data$cat12_tiv),] #remove cases with missing tiv values 
non_na_counts <- colSums(!is.na(tiv_data_subset)) #count the non missing values in each column 
tiv_data_subset <- tiv_data_subset[, non_na_counts > 0] #keep the columns which contain values 

#extract pseudonym from variable v1
tiv_data_subset$pseudonym <- as.character(tiv_data_subset$v1) #based on v1 create new variable "pseudondym" as character object 
tiv_data_subset$pseudonym <- gsub("^.{0,13}", "", tiv_data_subset$pseudonym) #remove charaters 0-13 
tiv_data_subset$pseudonym = substr(tiv_data_subset$pseudonym,1,nchar(tiv_data_subset$pseudonym)-22) #remove 22 tail characters
head(tiv_data_subset$pseudonym, 10) 
tiv_data_subset <- tiv_data_subset[, -1] #remove the original variable containing the pseudonyms 
```

PREPARATION SES DATA
```{r}
ses_data_subset <- ses_data[, c("Pseudonym", #choose subset of variables
                                "HCH_SVAGE0001", 
                                "HCH_SVSEX0001", 
                                "DEGS_HCHS_JOB_mean", 
                                "DEGS_HCHS_INCOME_mean", 
                                "DEGS_HCHS_EDU_mean",
                                "am_schube_ausb_037_2_ordinal_mean",
                                "HCHS_SES_Index_incl_imputed_data",
                                "vf_ace_001",
                                "vf_ace_002",
                                "vf_ace_003",
                                "vf_ace_004",
                                "vf_ace_005",
                                "vf_ace_006",
                                "vf_ace_007",
                                "vf_ace_008",
                                "vf_ace_009",
                                "vf_ace_010",
                                "vf_gad7_001",
                                "vf_gad7_002",
                                "vf_gad7_003",
                                "vf_gad7_004",
                                "vf_gad7_005",
                                "vf_gad7_006",
                                "vf_gad7_007",
                                "vf_phq9_001",
                                "vf_phq9_002",
                                "vf_phq9_003",
                                "vf_phq9_004",
                                "vf_phq9_005",
                                "vf_phq9_006",
                                "vf_phq9_007",
                                "vf_phq9_008",
                                "vf_phq9_009"
                                )]

colnames_ses_subset <- colnames(ses_data_subset) #save existing names to character vector
ses_cols_to_replace <- c("Pseudonym", "HCH_SVAGE0001", "HCH_SVSEX0001", "DEGS_HCHS_JOB_mean", "DEGS_HCHS_INCOME_mean", "DEGS_HCHS_EDU_mean", "am_schube_ausb_037_2_ordinal_mean", "HCHS_SES_Index_incl_imputed_data")
new_names <- c("pseudonym", "age", "sex", "job", "income", "education", "ind_income", "ses_index")
replace_indices <- which(colnames_ses_subset %in% ses_cols_to_replace)
colnames(ses_data_subset)[replace_indices] <- new_names
print(colnames(ses_data_subset))
```

PREPARE ACE VARIABLES 
```{r}
column_ace <- which(colnames(ses_data_subset) == "vf_ace_001") #Identify column of ace variables

for (i in column_ace:(column_ace + 9)) { # loop through columns containing ace variables
  print(unique(ses_data_subset[, i])) # check what kind of values exist --> result: some of the values include spaces
} 

vars_to_recode <- c("vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", "vf_ace_009", "vf_ace_010") #create a vector of variable names to recode 

for (i in 1:2) { # loop over selected variables and recode them (loop twice because otherwise there will be spaces left)
  for (var in vars_to_recode) {
    ses_data_subset[[var]] <- gsub("0 ", "0", ses_data_subset[[var]])
    ses_data_subset[[var]] <- gsub("1 ", "1", ses_data_subset[[var]])
    ses_data_subset[[var]] <- gsub("NA ", "NA", ses_data_subset[[var]])
    ses_data_subset[[var]] <- gsub("-99", "NA", ses_data_subset[[var]])
    ses_data_subset[[var]] <- gsub("NA$", "NA", ses_data_subset[[var]])  
    ses_data_subset[[var]][ses_data_subset[[var]] == "NA"] <- NA
  }
}

for (i in column_ace:(column_ace + 9)) { # loop through columns containing ace variables
  print(unique(ses_data_subset[, i])) # check if for loop succeeded 
} 
```
n(ses_data_subset)=10000 

CHECK FOR OUTLIERTS
```{r}
summary(ses_data_subset)

#create box plots for  selected variables
ses_boxplot <- c("job", "income", "education")
phq_boxplot <- c("vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", "vf_phq9_009")
gad_boxplot <- c("vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", "vf_gad7_005", "vf_gad7_006", "vf_gad7_007")
boxplot(ses_data_subset[, ses_boxplot])
boxplot(ses_data_subset[, phq_boxplot])
boxplot(ses_data_subset[, gad_boxplot])
```
Based on the boxplots, decided not to exclude outliers at this stage. Basically every response "3" on the PHQ-9 would be an outlier but we are interested in exactly these responses. Therefore, check for outliers after determination of sum scores again. 

IDENTIFY MISSING DATA (PATTERN)
```{r}
ses_missing_analysis <- ses_data_subset[, c("age", "sex", "job", "income", "education", "vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", "vf_ace_009", "vf_ace_010", "vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", "vf_gad7_005", "vf_gad7_006", "vf_gad7_007", "vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", "vf_phq9_009")]
ses_missing_analysis %>% #save variables of interest as new data frame and create a missing plot with them
  missing_plot(title = "Missing values plot of initial SES and questionnaire data (n = 10.000)")

explanatory = c("job", "income", "education", "vf_ace_001", "vf_gad7_001")
dependent = "vf_phq9_001"
ses_data_subset %>% # choose subset of variables and look at their missing data pattern
  missing_pattern(dependent, explanatory)

table_output <- miss_var_summary(ses_missing_analysis) # check for % of missing values per variable
table_html <- kable(table_output, format = "html") %>% # convert the table to a html table
  kable_styling()
writeLines(table_html, "missing_summary.html") #save html
```

MERGE MRI DATA SET + SES DATA SET + CAT12 TIV BY PSEUDONYM and CLEAN VARIABLE NAMES
```{r}
ses_data_subset$pseudonym <- trimws(ses_data_subset$pseudonym) # remove the spaces from the end of the pseudonym values
ace_mri_data <- merge(ses_data_subset, mri_data_complete, by = "pseudonym", all = TRUE) # merge data sets 
ace_mri_data <- merge(ace_mri_data, tiv_data_subset, by = "pseudonym", all = TRUE)

existing_names_preimp <- colnames(ace_mri_data) # check for present column names
columns_to_replace <- c("RightHippocampus", "RightAmygdala", "LeftHippocampus", "LeftAmygdala", "bilatHippocampus", "bilatAmygdala", "EstimatedTotalIntraCranialVol") 
new_names <- c("r_hippo", "r_amy", "l_hippo", "l_amy", "bilat_hippo", "bilat_amy", "tiv") # shorten long column names 
replace_indices <- which(existing_names_preimp %in% columns_to_replace)
colnames(ace_mri_data)[replace_indices] <- new_names
print(colnames(ace_mri_data)) # print to check out new column names
```
n=10004

REMOVE CASES THAT HAVE ONLY NAs ON MRI VARIABLES
```{r}
vars_mri <- c("r_hippo", "r_amy", "l_hippo", "l_amy", "bilat_hippo", "bilat_amy", "lhCortexVol", "rhCortexVol", "TotalGrayVol", "rh_G_front_middle_thickness", "rh_G_front_sup_thickness", "tiv")
vars_mri_na_only <- rowSums(is.na(ace_mri_data[, vars_mri])) == length(vars_mri) #identify cases that contain only NAs for vars_mri
ace_mri_data <- ace_mri_data %>%
  mutate(vars_mri_na_only = if_else(rowSums(is.na(select(., vars_mri))) == length(vars_mri), 1, 0)) %>%
  filter(vars_mri_na_only == 0) %>%
  select(-vars_mri_na_only)
```
n=2525

REMOVE CASES WITH ANOMALIES IN THE MRI AND MRI OUTLIERS 
```{r}
mri_clean <- readLines("mri_clean.txt") # load txt file with IDs of all cases that did not have any anomalies in the MRI, neither were they outlier
column_names <- c("pseudonym", "mri_clean")
mri_clean <- data.frame(pseudonym = mri_clean, mri_clean = 1) # create df with variable indicating that these mris are clean ("1"); these are n=2297
sum(is.na(ace_mri_data$pseudonym)) #check if there are NAs 
ace_mri_data <- merge(ace_mri_data, mri_clean, by = "pseudonym", all = TRUE) # merge dfs and keep all cases of both files
ace_mri_data$mri_clean[is.na(ace_mri_data$mri_clean)] <- 0
table(ace_mri_data$mri_clean) # n=327 cases have NAs and will therefore be excluded
ace_mri_data <- filter(ace_mri_data, mri_clean == 1) # keep all variables with mri_clean = 1, because these are the ones without amenities on mri --> n=2297
```
I have removed n=327 cases, because the MRIs did not pass the quality checks or because the IDs did match between the MRI file and the behavioral data file. --> n=2297

Create variables which indicate missing values for each case in each ses variable. The variables will be used in multiple imputation, however, these variables have been imputed in another study so we do not want to impute the missing values again but keep the NAs. Therefore, I save these NAs now, to later recreate them.
```{r}
check_na <- c("job", "education", "ind_income", "income") # Define the variables to check for NA values
for (variable in check_na) { # iterate over the variables 
  na_variable <- paste0(variable, "_na") # create a new variable with _na suffix for each variable
  ace_mri_data[[na_variable]] <- ifelse(is.na(ace_mri_data[[variable]]), 0, 1) # check for NAs in the variables and assign 0/1 accordingly
}
table(ace_mri_data$ind_income_na) # NAs n=102 
table(ace_mri_data$education_na) # NAs n=218 (same as income and job)
```

INSPECT MISSING DATA PATTERN IN QUESTIONNAIRE DATA  
To do so, remove those cases with NAs on the SES-Index from the data set because they won't be part of any analysis later.
```{r}
preimp_na_analysis <- ace_mri_data
preimp_na_analysis <- preimp_na_analysis[!is.na(preimp_na_analysis$ses_index), ] # select those cases that do not have an NA on SES-Index n=102
sum(is.na(preimp_na_analysis$ind_income))
sum(is.na(preimp_na_analysis$ses_index))
sum(is.na(ace_mri_data$ses_index))
```
dataset preimp_na_analysis (n=2195) is used to analyse missing data pattern in questionnaire data 

Create lists of variables belonging to each questionnaire to do separate NA analyses. 
```{r}
vars_ace <- c("vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", "vf_ace_009", "vf_ace_010") # save questionnaire items as list
vars_gad7 <- c("vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", "vf_gad7_005", "vf_gad7_006", "vf_gad7_007") # save the relevant questionnaire items as list
vars_phq9 <- c("vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", "vf_phq9_009") # save questionnaire items as list
vars_quest <- c("vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", "vf_ace_009", "vf_ace_010", "vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", "vf_gad7_005", "vf_gad7_006", "vf_gad7_007", "vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", "vf_phq9_009") # save all relevant questionnaire items as list

vis_miss(preimp_na_analysis[, vars_quest]) # visualize missing data pattern 

# For overview, create new variables for each questionnaire, indicating number of NAs per questionnaire item
preimp_na_analysis$ace_na_sum <- rowSums(is.na(preimp_na_analysis[, vars_ace])) # create new variable to store sum of ACE NAs
table(preimp_na_analysis$ace_na_sum) # check distribution of NAs
preimp_na_analysis$gad7_na_sum <- rowSums(is.na(preimp_na_analysis[, vars_gad7])) # create new variable to store sum GAD7 NAs
table(preimp_na_analysis$gad7_na_sum)
preimp_na_analysis$phq9_na_sum <- rowSums(is.na(preimp_na_analysis[, vars_phq9])) # create new variable to store sum of PHQ9 NAs
table(preimp_na_analysis$phq9_na_sum)
preimp_na_analysis$quest_na_sum <- rowSums(is.na(preimp_na_analysis[, vars_quest])) # create new variable to store sum of all questionnaire variables' NAs
print("quest_na_sum")
table(preimp_na_analysis$quest_na_sum)

#Remove those cases that have only NAs on either of the 3 questionnaires
vars_quest_na_only <- rowSums(is.na(preimp_na_analysis[, vars_quest])) == length(vars_quest) #identify cases that contain only NAs for vars_quest
preimp_na_analysis <- preimp_na_analysis %>%
  mutate(vars_quest_na_only = if_else(rowSums(is.na(select(., vars_quest))) == length(vars_quest), 1, 0))
table(preimp_na_analysis$vars_quest_na_only) # n=119

preimp_na_analysis <- preimp_na_analysis %>% 
  filter(vars_quest_na_only != 1) %>%
  select(-vars_quest_na_only) 

vars_phq9_na_only <- rowSums(is.na(preimp_na_analysis[, vars_phq9])) == length(vars_phq9) # identify cases that contain only NAs for vars_phq9
preimp_na_analysis <- preimp_na_analysis %>% #determine how many cases have NAs on all items of PHQ-9
  mutate(vars_phq9_na_only = if_else(rowSums(is.na(select(., vars_phq9))) == length(vars_phq9), 1, 0))
table(preimp_na_analysis$vars_phq9_na_only) # n=129 (n=10 after removal of those with only NAs on quest.)

preimp_na_analysis <- preimp_na_analysis %>%   
  filter(vars_phq9_na_only != 1) %>%
  select(-vars_phq9_na_only)

vars_gad7_na_only <- rowSums(is.na(preimp_na_analysis[, vars_gad7])) == length(vars_gad7) #identify cases that contain only NAs for vars_gad7
preimp_na_analysis <- preimp_na_analysis %>% #determine how many cases have NAs on all items of GAD-7
  mutate(vars_gad7_na_only = if_else(rowSums(is.na(select(., vars_gad7))) == length(vars_gad7), 1, 0))
table(preimp_na_analysis$vars_gad7_na_only) # n=131  (n=10 after removal of those with only NAs on quest.)

preimp_na_analysis <- preimp_na_analysis %>% 
  filter(vars_gad7_na_only != 1) %>%
  select(-vars_gad7_na_only)

vars_ace_na_only <- rowSums(is.na(preimp_na_analysis[, vars_ace])) == length(vars_ace) #identify cases that contain only NAs for vars_ace
preimp_na_analysis <- preimp_na_analysis %>%
  mutate(vars_ace_na_only = if_else(rowSums(is.na(select(., vars_ace))) == length(vars_ace), 1, 0))
table(preimp_na_analysis$vars_ace_na_only) # n=134  (n=14 after removal of those with only NAs on quest.)

preimp_na_analysis <- preimp_na_analysis %>% 
  filter(vars_ace_na_only != 1) %>%
  select(-vars_ace_na_only)

vis_miss(preimp_na_analysis[, vars_quest]) #visualize missing data pattern
table(preimp_na_analysis$ace_na_sum)
table(preimp_na_analysis$gad7_na_sum)
table(preimp_na_analysis$phq9_na_sum)
table(preimp_na_analysis$quest_na_sum)
colSums(is.na(preimp_na_analysis[, vars_quest]))
```
n(preimp_na_analysis)=2042

MAIN DATASET (PRE-IMPUTATION DATASET)
Determine QUESTTIONNAIRE SUM SCORES of non-imputed data + remove all NAs to proceed with mediation analysis
```{r}
preimp_no_na <- preimp_na_analysis[, c( # create new data frame containing relevant variables 
                                        "pseudonym", 
                                        "age", 
                                        "sex", 
                                        "ses_index", 
                                        "vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", 
                                        "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", 
                                        "vf_ace_009", "vf_ace_010", 
                                        "vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", 
                                        "vf_gad7_005", "vf_gad7_006", "vf_gad7_007", 
                                        "vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", 
                                        "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", 
                                        "vf_phq9_009", 
                                        "r_hippo", "r_amy", "l_hippo", "l_amy", 
                                        "bilat_hippo", "bilat_amy", 
                                        "lhCortexVol", "rhCortexVol", "TotalGrayVol", 
                                        "rh_G_front_middle_thickness", "rh_G_front_sup_thickness", 
                                        "tiv", "cat12_tiv"
                                        )] 
preimp_no_na <- mutate_if(preimp_no_na, function(x) !is.character(x), as.numeric) # for next step needed to convert to numeric
sapply(preimp_no_na, class)

# ACE: determine sum score
preimp_no_na$ace_pre_imp <- rowSums(preimp_no_na[, vars_ace] == 1, na.rm = TRUE)
table(preimp_no_na$ace_pre_imp)
has_missing_ace <- apply(preimp_no_na[, vars_ace], 1, function(x) any(is.na(x))) # Check for NA
preimp_no_na$ace_pre_imp[has_missing_ace] <- NA # Replace sum score with NA if there are missing values
table(preimp_no_na$ace_pre_imp)
table(has_missing_ace) # n=83

#PHQ-9: determine sum score
preimp_no_na$phq9_pre_imp <- rowSums(preimp_no_na[, vars_phq9], na.rm = TRUE) 
table(preimp_no_na$phq9_pre_imp)
has_missing_phq <- apply(preimp_no_na[, vars_phq9], 1, function(x) any(is.na(x)))
preimp_no_na$phq9_pre_imp[has_missing_phq] <- NA
table(preimp_no_na$phq9_pre_imp)
table(has_missing_phq) # n=48

#GAD-7: determine sum score
preimp_no_na$gad7_pre_imp <- rowSums(preimp_no_na[, vars_gad7], na.rm = TRUE)
table(preimp_no_na$gad7_pre_imp)
has_missing_gad <- apply(preimp_no_na[, vars_gad7], 1, function(x) any(is.na(x)))
preimp_no_na$gad7_pre_imp[has_missing_gad] <- NA
table(preimp_no_na$gad7_pre_imp)
table(has_missing_gad) # n=25

preimp_no_na <- preimp_no_na[complete.cases(preimp_no_na[c("ace_pre_imp", "phq9_pre_imp", "gad7_pre_imp")]), ] #remove NAs from dataset for named variables

hist(preimp_no_na$ace_pre_imp, main = "Histogram of ACE sum score after exclusion of NAs, 
before imputation, before exclusion of outliers (n = 1900)")
hist(preimp_no_na$phq9_pre_imp, main = "Histogram of PHQ9 sum score after exclusion of NAs, 
before imputation, before exclusion of outliers (n = 1900)")
hist(preimp_no_na$gad7_pre_imp, main = "Histogram of GAD7 sum score after exclusion of NAs, 
before imputation, before exclusion of outliers (n = 1900)")

sum(is.na(preimp_no_na))
```
preimp_no_na: n=1900

DETERMINE DLPFC VALUE BASED ON MEAN OF middle frontal gyrus and superior frontal gyrus
```{r}
preimp_no_na$dlpfc <- (preimp_no_na$rh_G_front_middle_thickness + preimp_no_na$rh_G_front_sup_thickness) / 2
```

CREATE AN ORDINAL AND DICHOTOMOUS VARIABLE FOR ACE (0, 1, 2, 3, >=4) AND (0, 1)
```{r}
# dichotomous
preimp_no_na$ace_pre_imp_dich <- ifelse(preimp_no_na$ace_pre_imp == 0, 0, 1) #create dichotomous variable

# ordinal 0-4
breaks <- c(-Inf, 0, 1, 2, 3, 10) # define the breaks for cut-off points of the ordinal categories
preimp_no_na$ace_pre_imp_ord_4 <- cut(preimp_no_na$ace_pre_imp, breaks = breaks, #using cut, convert numeric variable into ordinal; using breaks define cut-offs;  
                                    labels = c(0, 1, 2, 3, 4), include.lowest = TRUE) #using labels determine labels; using include.lowest include lower bound of first category
table(preimp_no_na$ace_pre_imp) # check if it was done correctly
table(preimp_no_na$ace_pre_imp_ord_4)

# categories 0, 1-3, 4-10
breaks <- c(-Inf, 0, 3, 10) # define the breaks for cut-off points of the ordinal categories
preimp_no_na$ace_pre_imp_categ <- cut(preimp_no_na$ace_pre_imp, breaks = breaks, # using cut, convert numeric variable into ordinal; using breaks define cut-offs;  
                                    labels = c(0, 1, 2), include.lowest = TRUE) # using labels determine labels; using include.lowest include lower bound of first category
table(preimp_no_na$ace_pre_imp) # check if it was done correctly
table(preimp_no_na$ace_pre_imp_categ)
```

Create an ordinally structured ace variable for non-monotonic relationship test in jasp
```{r}
breaks <- c(-Inf, 0, 1, 2, 3, 4, 5, 6, 10) #define the breaks for cut-off points of the ordinal categories
preimp_no_na$ace_pre_imp_ord_7 <- cut(preimp_no_na$ace_pre_imp, breaks = breaks, #using cut, convert numeric variable into ordinal; using breaks define cut-offs;  
                                    labels = c(0, 1, 2, 3, 4, 5, 6, 7), include.lowest = TRUE) #using labels determine labels; using include.lowest include lower bound of first category

table(preimp_no_na$ace_pre_imp) #check if it was done correctly
table(preimp_no_na$ace_pre_imp_ord_7)
```

CHECK FOR OUTLIERS
```{r}
variables_of_interest <- c("ace_pre_imp", "phq9_pre_imp", "gad7_pre_imp") # specify the variables of interest

outliers <- logical(nrow(preimp_no_na)) # create empty logical vector to store outlier flags

for (variable in variables_of_interest) { # calculate mean and standard deviation for the current variable
  variable_mean <- mean(preimp_no_na[[variable]], na.rm = TRUE)
  variable_sd <- sd(preimp_no_na[[variable]], na.rm = TRUE)  # Identify data points outside the specified range for the current variable
  variable_outliers <- preimp_no_na[[variable]] < (variable_mean - 3 * variable_sd) | preimp_no_na[[variable]] > (variable_mean + 3 * variable_sd) # Update the outlier flags
  outliers <- outliers | variable_outliers
}

outlier_data <- preimp_no_na[outliers, ] # Subset the dataset to obtain the outliers

preimp_excl_outlier <- preimp_no_na[!(rownames(preimp_no_na) %in% rownames(outlier_data)), ] # exclude cases in outlier_data from preimp_no_na

preimp_excl_outlier$dlpfc <- (preimp_excl_outlier$rh_G_front_middle_thickness + preimp_excl_outlier$rh_G_front_sup_thickness) / 2
```
n(preimp_excl_outlier)=1815
n(ace_mri_data)=2297

PREPARE DATASET FOR IMPUTATION: REMOVE CASES WITH ONLY MISSING VALUES ON SPECIFIC VARIABLES
EXCLUDE THOSE CASES THAT HAVE MISSING VALUES ON ALL OF THE SES VARIABLES OR ON ALL OF THE QUESTIONNAIRE ITEMS 
```{r}
vars_quest_na_only <- rowSums(is.na(ace_mri_data[, vars_quest])) == length(vars_quest) # identify cases that contain only NAs for the questionnaire variables 

ses_data_subset_excl_na <- ace_mri_data %>%
  mutate(vars_quest_na_only = if_else(rowSums(is.na(select(., vars_quest))) == length(vars_quest), 1, 0)) %>% #create a new variable "vars_quest_na_only" which identifies those cases that have only NAs on the variables in "vars_quest"
  filter(vars_quest_na_only == 0) %>% #remove those with 1 on the newly created variable, n(removed)=221
  select(-vars_quest_na_only) # remove the auxiliary variable 

#Compare missing data pattern to before
ses_missing_analysis_excl_na <- ses_data_subset_excl_na[, c("age", "sex", "job", "income", "education", "vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", "vf_ace_009", "vf_ace_010", "vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", "vf_gad7_005", "vf_gad7_006", "vf_gad7_007", "vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", "vf_phq9_009")]
ses_missing_analysis_excl_na %>% 
  missing_plot() #missing values map for overview

vars_phq9_na_only <- rowSums(is.na(ses_data_subset_excl_na[, vars_phq9])) == length(vars_phq9) #identify cases that contain only NAs for vars_phq9
ses_data_subset_excl_na <- ses_data_subset_excl_na %>% #determine how many cases have NAs on all items of PHQ-9
  mutate(vars_phq9_na_only = if_else(rowSums(is.na(select(., vars_phq9))) == length(vars_phq9), 1, 0))
table(ses_data_subset_excl_na$vars_phq9_na_only) # n=10

vars_gad7_na_only <- rowSums(is.na(ses_data_subset_excl_na[, vars_gad7])) == length(vars_gad7) #identify cases that contain only NAs for vars_gad7
ses_data_subset_excl_na <- ses_data_subset_excl_na %>% #determine how many cases have NAs on all items of GAD-7
  mutate(vars_gad7_na_only = if_else(rowSums(is.na(select(., vars_gad7))) == length(vars_gad7), 1, 0))
table(ses_data_subset_excl_na$vars_gad7_na_only) # n=12

ses_data_subset_excl_na_quest_phq_gad <- ses_data_subset_excl_na %>%
  filter(vars_phq9_na_only == 0) %>% # n=10
  filter(vars_gad7_na_only == 0) %>% # n=12 after removal of phq9 variables
  select(-vars_phq9_na_only) %>%
  select(-vars_gad7_na_only) # n(removed)=20

ses_missing_analysis_excl_na_quest_phq_gad <- ses_data_subset_excl_na_quest_phq_gad[, c("age", "sex", "job", "income", "education", "vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", "vf_ace_009", "vf_ace_010", "vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", "vf_gad7_005", "vf_gad7_006", "vf_gad7_007", "vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", "vf_phq9_009")] #Compare missing data pattern to before
ses_missing_analysis_excl_na_quest_phq_gad %>% 
  missing_plot() #missing values map for overview

vars_ace_na_only <- rowSums(is.na(ses_data_subset_excl_na_quest_phq_gad[, vars_ace])) == length(vars_ace) #identify cases that contain only NAs for vars_ace
table(vars_ace_na_only) # n(only NA)=14
ses_data_subset_excl_na_quest_phq_gad_ace <- ses_data_subset_excl_na_quest_phq_gad %>%
  mutate(vars_ace_na_only = if_else(rowSums(is.na(select(., vars_ace))) == length(vars_ace), 1, 0)) %>%
  filter(vars_ace_na_only == 0) %>%
  select(-vars_ace_na_only) # n(removed)=14
ace_mri_preimp <- ses_data_subset_excl_na_quest_phq_gad_ace

ses_missing_analysis_excl_na_quest_phq_gad_ace <- ace_mri_preimp[, c("age", "sex", "job", "income", "education", "vf_ace_001", "vf_ace_002", "vf_ace_003", "vf_ace_004", "vf_ace_005", "vf_ace_006", "vf_ace_007", "vf_ace_008", "vf_ace_009", "vf_ace_010", "vf_gad7_001", "vf_gad7_002", "vf_gad7_003", "vf_gad7_004", "vf_gad7_005", "vf_gad7_006", "vf_gad7_007", "vf_phq9_001", "vf_phq9_002", "vf_phq9_003", "vf_phq9_004", "vf_phq9_005", "vf_phq9_006", "vf_phq9_007","vf_phq9_008", "vf_phq9_009")]
ses_missing_analysis_excl_na_quest_phq_gad_ace %>% 
  missing_plot(title = "Missing values plot of SES and questionnaire data (n=2.042)") 
```
09/07/2023 + 06/03/2024 ace_mri_preimp n=2042

ADAPTATION DATA TYPES 
```{r}
str(ace_mri_preimp)# check current data type
sum(is.na(ace_mri_preimp))

for (i in  9:18) {
  ace_mri_preimp[, i] <- as.logical(as.integer(ace_mri_preimp[, i]))
}
ace_mri_preimp$sex <- as.logical(ace_mri_preimp$sex)

for (i in 26:34) { # change type of PHQ variables from numeric to categorical (ordered) (necessary for correct imputation method)
  ace_mri_preimp[[i]] <- cut(ace_mri_preimp[[i]], 
                             breaks = 4,
                             labels = c("0", "1", "2", "3"), 
                             ordered = TRUE)
  names(ace_mri_preimp)[i] <- paste0(names(ace_mri_preimp)[i])
}

for (i in 19:25) { # change type of GAD variables from numeric to categorical (ordered) (necessary for correct imputation method)
  ace_mri_preimp[[i]] <- cut(ace_mri_preimp[[i]], 
                             breaks = 4,
                             labels = c("0", "1", "2", "3"), 
                             ordered = TRUE)
  names(ace_mri_preimp)[i] <- paste0(names(ace_mri_preimp)[i])
}

ace_mri_preimp <- ace_mri_preimp %>% #declare NAs as NAs because somehow some of them have been saved as values
  mutate(
    job = replace(job, job == "NA", NA),
    income = replace(income, income == "NA", NA),
    education = replace(education, education == "NA", NA)
  )

sum(is.na(ace_mri_preimp$education))
sum(is.na(ace_mri_preimp))
```
09/07/2023 + 06/03/2024 Remaining NAs in SES variables n=92. Total remaining NAs n=629.

Save as csv files
```{r}
# Save preimp_no_na.csv dataframe (earlier ace_pre_imp) which will be used for analyses 
write.csv(preimp_no_na, file = "preimp_no_na_1900.csv", row.names = FALSE)

# Save preimp_no_na_excl_outlier.csv dataframe which will be used for sensitivity analyses 
write.csv(preimp_excl_outlier, file = "preimp_no_na_excl_outlier_1815.csv", row.names = FALSE)

# Save preimp_prepped_for_imputation dataframe which will be used to impute missing data in "02_imputation.Rmd"
write.csv(ace_mri_preimp, file = "preimp_prepped_for_imputation_2042.csv", row.names = FALSE)

#Save pseudonym values of all cases in the n=1900 sample
length(preimp_no_na$pseudonym) # check for number of cases 
sum(is.na(preimp_no_na$pseudonym)) # check for NAs
pseudonym_values_1900 <- preimp_no_na$pseudonym
values_string_1900 <- paste(pseudonym_values_1900, collapse = "\n") # Create a string with values separated by newline character
write(values_string_1900, "pseudonym_values_preimp_no_na_1900.txt")  # Write the string to a text file

#Save pseudonym values of all cases in the n=1900 sample
length(preimp_excl_outlier$pseudonym) # check for number of cases 
sum(is.na(preimp_excl_outlier$pseudonym)) # check for NAs
pseudonym_values_1815 <- preimp_excl_outlier$pseudonym
values_string_1815 <- paste(pseudonym_values_1815, collapse = "\n") # Create a string with values separated by newline character
write(values_string_1815, "pseudonym_values_preimp_no_na_excl_outl_1815.txt")  # Write the string to a text file
```
