---
title: "2024_02_main_hypothesis_non_monotonic"
output: html_document
date: "2024-02-05"
author: A.Klimesch
references: Datacamp course "Generalized Linear Models in R"; OpenAI. (2023). ChatGPT (February 2024 version) [Large language model]. https://chat.openai.com/chat
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(tidyverse)
library(ggplot2)
library(GGally)  # for exploring relationships between pairs of variables
library(mgcv) 
library(gridExtra) # to save combined plots
```

Load dataset
```{r}
preimp_no_na <- read.csv("C:path_to_data/preimp_no_na_1900.csv")
```

CHECK NON-MONOTONIC RELATIONSHIP FOR preimp_no_na DATASET
Scatterplots preimp_no_na
```{r}
plot_relationship <- function(data, x_var, y_var, x_label = x_var, y_label = y_var, title = "Scatterplot") {
  p <- data %>%
    ggplot(aes_string(x = x_var, y = y_var)) +
    geom_jitter(width = 0.2, height = 0.2) +
    geom_smooth(method = "lm", col = "red") +
    geom_smooth(method = "loess", col = "blue") +
    labs(title = title, x = x_label, y = y_label)
  
  # Adjust for specific combinations
  if (y_var %in% c("phq9_pre_imp", "gad7_pre_imp")) {
    p <- p + geom_smooth(method = "glm", method.args = list(family = "poisson"), col = "yellow")
  }
  
  # Generate filename based on x_var and y_var
  filename <- paste(x_var, y_var, "scatterplot.png", sep = "_")
  full_path <- file.path(save_path, filename)
  
  # Save the plot
  ggsave(full_path, plot = p, width = 8, height = 6)
}

# Create a folder for the plots if it doesn't already exist
save_path <- "C:/your_path/scatterplots"
if (!dir.exists(save_path)) {
  dir.create(save_path)
}

# Define your variable combinations
x_vars <- c("ace_pre_imp_ord_7", "phq9_pre_imp", "gad7_pre_imp")
y_vars <- c("bilat_hippo", "dlpfc", "bilat_amy", "phq9_pre_imp", "gad7_pre_imp")

# Loop through each combination and plot
for (x in x_vars) {
  for (y in y_vars) {
    plot_relationship(preimp_no_na, x, y)
  }
}
```

Check for normality of residuals of continuous variables (ChatGPT 4 was used)
```{r}
p1 <- ggplot(preimp_no_na,aes(sample=bilat_amy)) + 
  geom_qq() + 
  geom_qq_line() + 
  ggtitle('Amygdala')

p2 <- ggplot(preimp_no_na,aes(sample=dlpfc)) + 
  geom_qq() + 
  geom_qq_line()+ 
  ggtitle('DLPFC')

p3 <- ggplot(preimp_no_na,aes(sample=bilat_hippo)) + 
  geom_qq() + 
  geom_qq_line() + 
  ggtitle('Hippo')

combined_plot <- arrangeGrob(p1, p2, p3, nrow = 1) # Combine plots without displaying them
plots_directory <- "C:/your_path/normality_plot" # define directory

if (!dir.exists(plots_directory)) { # Check if the directory exists, if not, create it
  dir.create(plots_directory)
}

plot_filename <- "check_for_normality_of_residuals.png" # Define a unique filename for the new plot
save_path <- file.path(plots_directory, plot_filename) # Full path to save the new plot
ggsave(file = save_path, plot = combined_plot, width = 20, height = 7, units = "cm") # Save the plot
```

Poisson regression for behavioral variables (using datacamp code)
```{r}
poisson_phq9 <- glm(phq9_pre_imp ~ ace_pre_imp_ord_7, data = preimp_no_na, family = "poisson") # fit y predicted by x with data.frame preimp_no_na using the poisson family
print(poisson_phq9)

poisson_gad7 <- glm(gad7_pre_imp ~ ace_pre_imp_ord_7, data = preimp_no_na, family = "poisson")
print(poisson_gad7)

coef(poisson_phq9) # Extract the regression coefficients
coef(poisson_gad7)
confint(poisson_phq9) # Extract the confidence intervals
confint(poisson_gad7)
summary(poisson_phq9) # Examine the outputs using summary
summary(poisson_gad7)
```

For comparison: Linear regression with gaussian distribution for neuro variables (using datacamp code)
```{r}
gaussian_dlpfc <- glm(dlpfc ~ ace_pre_imp_ord_7, data = preimp_no_na, family = "gaussian") # fit y predicted by x with data.frame preimp_no_na using the poisson family
print(gaussian_dlpfc)
summary(gaussian_dlpfc)

gaussian_hippo <- glm(bilat_hippo ~ ace_pre_imp_ord_7, data = preimp_no_na, family = "gaussian")
print(gaussian_hippo)
summary(gaussian_hippo)

gaussian_amy <- glm(bilat_amy ~ ace_pre_imp_ord_7, data = preimp_no_na, family = "gaussian")
print(gaussian_amy)
summary(gaussian_amy)
```

Perform several generalized additive models and create their pdp plots to validate the results on non-monotonic relationships (ChatGPT 4 was used)
```{r}
output_dir <- "C:/your_path/pdp_plots"

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

variable_pairs <- list( # define your variable pairs and covariates
  c("ace_pre_imp_ord_7", "bilat_amy"),
  c("ace_pre_imp_ord_7", "bilat_hippo"),
  c("ace_pre_imp_ord_7", "dlpfc"),
  c("bilat_hippo", "phq9_pre_imp"),
  c("bilat_amy", "phq9_pre_imp"),
  c("dlpfc", "phq9_pre_imp"),
  c("bilat_hippo", "gad7_pre_imp"),
  c("bilat_amy", "gad7_pre_imp"),
  c("dlpfc", "gad7_pre_imp")
)

for (pair in variable_pairs) { # Loop through each pair, fit a GAM, and print the summary
  response <- pair[2]
  predictor <- pair[1]
  
  formula <- as.formula(paste(response, "~ s(", predictor, ", k=5) + age + sex", sep = "")) # formula with specified 'k' value to limit the complexity of the smooth
  
  gam_model <- gam(formula, data = preimp_no_na, method = "REML") # Fit the GAM model
  
  filename <- sprintf("pdp_of_gam_preimp_%s_vs_%s.png", predictor, response)
  full_path <- file.path(output_dir, filename)
  
  png(file = full_path, width = 800, height = 600) # You can adjust the size as needed
  
  plot(gam_model, select = 1, main = paste("PDP for", predictor, "on", response))
  dev.off()
  
  print(summary(gam_model)) # Print the summary of the model and plot
}
```
The only significant GAM is the one with x = bilat_amy, y = gad7_pre_imp BUT their plot shows that the CIs cross 0, thus the significant result only holds for certain ranges of bilat_amy and we cannt conclude a non-monotonic relationship between x and y with certainty.  

